---
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Directory

# Load Libraries
```{r load libraries, include=FALSE}
library(tidyverse)
library(ggplot2)
library(MASS)
library(gridExtra)
```

# Loading the dataset and check its structure
```{r include=FALSE}
dat <- Boston
str(dat) #check data structure
summary(dat) #check the spread of each variable
head(dat) 
tail(dat)
#no extraneous values have been read in
```
##check data structure
##check the spread of each variable

# Data Preparation
```{r include=FALSE}
sum(is.na(dat))#Check for NA values - 0
#Inspect the skewness for variables with a larger range
hist(dat$crim) #right skewed
hist(dat$zn) #right skewed
hist(dat$indus) #right skewed
hist(dat$nox) #right skewed
hist(dat$rm) #right skewed
hist(dat$age) #left skewed 
hist(dat$dis) #right skewed
hist(dat$rad) #right skewed
hist(dat$tax) #right skewed
hist(dat$ptratio) #left skewed
hist(dat$black) #left skewed
hist(dat$lstat) #right skewed
hist(dat$medv) #right skewed
```
##Inspect the skewness for variables with a larger range


## Objective: Build regression models to explain real estate price variation.
## in this case it's the "medv" being the dependent variable.

## Plot against variables
```{r include=FALSE}
#medv vs crim
ggplot(dat, aes(x=crim, y=medv)) + geom_point() + geom_smooth() + scale_x_log10()
ggplot(dat, aes(x=crim, y=medv)) + geom_point() + geom_smooth()
##After log x transformation, see a smoother curve with an explainable trend. As crime rate increase, the median value of owner-occupied homes fall.

#medv vs zn
ggplot(dat, aes(x=zn, y=medv)) + geom_point() + geom_smooth()
ggplot(dat, aes(x=zn, y=medv)) + geom_jitter(alpha = 0.4) + geom_smooth() #to avoid over-crowding of points
##As the proportion of residential land zoned for lots over 25,000 sq.ft (zn) rises, the median-value of owner-occupied homes rises as well which is expected since most of the houses at higher zn should be the bigger houses that would naturally have a higher median value.

#medv vs indus
ggplot(dat, aes(x=indus, y=medv)) + geom_point(position = "jitter", alpha = 0.4) + geom_smooth()
##As the proportion of non-retail business acres per town rises, the median value of owner-occupied homes falls.
##This make sense because in areas where there are higher retail shops, the mdev is higher and vice-versa.

#medv vs chas
#ggplot(dat, aes(x=chas, y=medv)) + geom_point() + geom_smooth()
## Since this is a dummy variable, cannot do a ggplot for this 
ggplot(dat, aes(x=chas, y=medv, group=chas)) + geom_boxplot() + geom_jitter(alpha = 0.4, col = "orange")
##The median value of houses are higher where the tract bounds the Charles river.
##Otherwise, the medv of the houses are lower. 

#medv vs nox
ggplot(dat, aes(x=nox, y=medv)) + geom_point(position = "jitter", alpha = 0.4) + geom_smooth()
##Not a very smooth line but at least there is an expected downward trend where median value of houses are lower as the
#concentration of nitric oxide increases

#medv vs rm
ggplot(dat, aes(x=rm, y=medv)) + geom_point(position = "jitter", alpha = 0.4) + geom_smooth()
##There is a positive trend and that is expected because bigger houses 
##would fetch higher median values. However, there are some outliers.

#medv vs age
ggplot(dat, aes(x=age, y=medv)) + geom_point(position = "jitter", alpha = 0.4) + geom_smooth()
##Same trend as nox but this is only because of the large concentration of many old houses at the tail end.

#medv vs dis
ggplot(dat, aes(x=dis, y=medv)) + geom_point(position = "jitter", alpha = 0.4) + geom_smooth()
##Generally, there is no correlation between the medv and the weighted distances to five Boston employment centres.
#However, for housing too close to the employment centres (~ < 3km), the medv can be observed to be lower.

#medv vs rad
ggplot(dat, aes(x=rad, y=medv)) + geom_point() + geom_smooth()
ggplot(dat, aes(x=rad, y=medv)) + geom_boxplot(aes(group=rad)) + geom_jitter(alpha = 0.4, col = "orange")
##There is no discernible trend. However, the skewness of the medv of homes increases (1-3) & (4-8).
#This suggests that the accessibility of homes to highways raises the value of homes. 
#outlier is when the radial index of 24 highways observed a sharp fall in medv. 

#medv vs ptratio
ggplot(dat, aes(x=ptratio, y=medv)) + geom_jitter(alpha = 0.4) + geom_smooth()
##Again, downward trend is observed where higher ptratio results in lower house prices. 

#medv vs black
ggplot(dat, aes(x=black, y=medv)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
##There is not a clear trend between the proportion of black population by town and the median house values.
##However, when the value of black=400, a majority of medv of houses cluster at the lower end of the medv values. 

#medv vs lstat
ggplot(dat, aes(x=lstat, y=medv)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
##Downward trend where the median price is lower at high levels of lstat (lower status of the population). 
```

## Plot against variables
##Plots were done for all variables but only showing against rm as an example
```{r echo=TRUE}
#medv vs rm
ggplot(dat, aes(x=rm, y=medv)) + geom_point(position = "jitter", alpha = 0.4) + geom_smooth()
##There is a positive trend and that is expected because bigger houses 
##would fetch higher median values. However, there are some outliers.
```


## Fitting a linear Model rm, lstat, nox, age, ptratio

```{r echo=TRUE}
lm1 <- lm(medv ~ crim + indus + rm + tax + lstat + nox + age + ptratio, data=dat)
lm1
##Most of the sign of the coefficients make sense except age where the relationship was thought to be negative.
lm1_summary <- summary(lm1)
lm1_summary
##Only rm, lstat, ptratio are significant at the 0.001 level. age is 
##significant at 0.05 level. nox is also significant at the 0.1 level.

##Extracting R2, RMSPE, and AIC
r2a <- lm1_summary$adj.r.squared
r2a #0.6811
rmspe1 <- sqrt(mean(rstandard(lm1, type="predictive")^2)) 
rmspe1 #5.295
AIC(lm1) #3114.086

##Fitting a model and excluding non-significant variables 

lm2 <- lm(medv ~ rm + lstat + nox + age + ptratio , data=dat)
lm2
##Most of the sign of the coefficients make sense except age where the relationship is expected to be negative.
lm2_summary <- summary(lm2)
lm2_summary

# Extracting R2, RMSPE, and AIC
r2b <- lm2_summary$adj.r.squared 
r2b #0.680
rmspe2 <- sqrt(mean(rstandard(lm2, type="predictive")^2)) 
rmspe2 #5.278
AIC(lm2) #3112.307

#AIC(lm2) < AIC(lm1)
#rmspe2 < rmspe1
#r2b < r2a
```

## Default plots from R

```{r include=FALSE}
plot(lm2,which=1:5)
#Influential points: 215, 366, 369, 372, 373
#Plot 2: Histogram of residuals is right-skewed
#Plot 5: Most of the residual points have low residual value and leverage.
```
#These plots include the Residuals v fitted, Normal Q-Q, Scale-location, Cook's
#distance & Residuals vs leverage. 
#Influential points: 215, 366, 369, 372, 373
#Plot 2: Histogram of residuals is right-skewed
#Plot 5: Most of the residual points have low residual value and leverage.


```{r echo=TRUE}
# Adding new columns to the existing dataset
dat <- mutate(dat, nstud_r = resid(lm2), stud_r = studres(lm2), 
              fit_lm2 = fitted(lm2))

# Create residual plots for each variable

ggplot(dat) + geom_histogram(aes(x=stud_r))
##based on stud_r histogram, indeed residuals are right skewed with outliers

p1 <- ggplot(dat,aes(x=rm, y=medv)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
p2 <- ggplot(dat,aes(x=rm, y=stud_r)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
p3 <- ggplot(dat,aes(x=lstat, y=medv)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
p4 <- ggplot(dat,aes(x=lstat, y=stud_r)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
p5 <- ggplot(dat,aes(x=nox, y=medv)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
p6 <- ggplot(dat,aes(x=nox, y=stud_r)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
p7 <- ggplot(dat,aes(x=age, y=medv)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
p8 <- ggplot(dat,aes(x=age, y=stud_r)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
p9 <- ggplot(dat,aes(x=ptratio, y=medv)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
p10 <- ggplot(dat,aes(x=ptratio, y=stud_r)) + geom_point(position="jitter", alpha=0.4) + geom_smooth()
grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10, ncol=2) #Plots of x and y variables and residual plots
#of each of the model's variables

##Based on the above grid.arrange plot, the residuals for all the 
##chosen variables are clustered around 0. 

#p1 <- ggplot(dat) + geom_boxplot(aes(x=rm, y=stud_r, group=rm))
#p2 <- ggplot(dat) + geom_boxplot(aes(x=rm, y=medv, group=rm))
#grid.arrange(p1, p2, nrow=2)
```


## Detecting influential points
```{r include=FALSE}
#Output of dfbetas has 6 columns - 1 for each explanatory variable including the intercept 
dfb <- dfbetas(lm2)
qplot(x=1:506, y = dfb[,1]) #influential points on the intercept beta
which(dfb[,1] > 0.3) #365 366 368 369
qplot(x=1:506, y = dfb[,2]) #influential points on the rm beta
which(dfb[,2] < -0.4) #365 366 368 369
qplot(x=1:506, y = dfb[,3]) #influential points on the lstat beta
which(dfb[,3] < -0.75 |dfb[,3] > 0.5) #215 369
qplot(x=1:506, y = dfb[,4]) #influential points on the nox beta
which(dfb[,4] < -0.2|dfb[,4] > 0.2) #156 365 366 373 413 
qplot(x=1:506, y = dfb[,5]) #influential points on the age beta
which(dfb[,5] < -0.45|dfb[,5] > 0.45) #215 369 
qplot(x=1:506, y = dfb[,6]) #influential points on the ptratio beta
which(dfb[,6] < -0.2|dfb[,6] > 0.2) #365 370 371 372 373
#From the plots, likely influential points are: 365 369
```
##From the plots, likely influential points are: 365 369


## Method 1: Fitting the robust line
```{r echo=TRUE}
lm2 <- lm(medv ~ rm + lstat + nox + age + ptratio, data=dat)
lm2
rlm2 <- rlm(medv ~ rm + lstat + nox + age + ptratio, data=dat) #robust linear model
rlm2

#Using values of slope for each coefficient to fit the robust lines.

#fitting rm and medv
fitted_lm1 <- ggplot(dat) + geom_point(aes(x=rm,y=medv)) +
  geom_smooth(aes(x=rm,y=medv),method="lm",se=FALSE)
fitted_rlm1 <- fitted_lm1 + geom_abline(slope=5.15770961,intercept=16.46533520,col="red")

#fitting lstat and medv
fitted_lm2 <- ggplot(dat) + geom_point(aes(x=lstat,y=medv)) +
  geom_smooth(aes(x=lstat,y=medv),method="lm",se=FALSE)
fitted_rlm2 <- fitted_lm2 + geom_abline(slope=-0.45586584,intercept=16.46533520,col="red")

#fitting nox and medv
fitted_lm3 <- ggplot(dat) + geom_point(aes(x=nox,y=medv)) +
  geom_smooth(aes(x=nox,y=medv),method="lm",se=FALSE)
fitted_rlm3 <- fitted_lm3 + geom_abline(slope=-7.85963756,intercept=16.46533520,col="red")

#fitting age and medv
fitted_lm4 <- ggplot(dat) + geom_point(aes(x=age,y=medv)) +
  geom_smooth(aes(x=age,y=medv),method="lm",se=FALSE)
fitted_rlm4 <- fitted_lm4 + geom_abline(slope=0.01124983,intercept=16.46533520,col="red")

#fitting ptratio and medv
fitted_lm5 <- ggplot(dat) + geom_point(aes(x=ptratio,y=medv)) +
  geom_smooth(aes(x=ptratio,y=medv),method="lm",se=FALSE)
fitted_rlm5 <- fitted_lm5 + geom_abline(slope=-0.95109494,intercept=16.46533520,col="red")

#Putting them all together
grid.arrange(fitted_rlm1, fitted_rlm2, fitted_rlm3, fitted_rlm4, fitted_rlm5, nrow=2)
```


## Method 2: Dropping identified influential points
```{r echo=TRUE}
#Include id column to remove the influential points
dat <- mutate(dat, id = 1:nrow(dat))

#Filter data without influential points
dat_v2 <- dat %>%
  filter(id != 365 & id != 369) 
```

## An improved model
```{r echo=TRUE}
##Extracting R2, RMSPE, and AIC
lm3 <- lm(medv ~ rm + lstat + nox + age + ptratio, data=dat_v2)
lm3
lm3_summary <- summary(lm3)
r2c <- lm3_summary$adj.r.squared
r2c #0.700
rmspe3 <- sqrt(mean(rstandard(lm3, type="predictive")^2)) 
rmspe3 #5.068
AIC(lm3) #3060.546

#AIC is lower than initial model. 
#RMSPE is lower at 5.068. Initial: sd(dat_v2$medv): 9.133429
#RMSPE, AIC, R2 are all lower - the model with the omitted influential points should be used. 

